\documentclass[11pt]{article}
\usepackage[margin=2.6cm]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{amsmath,amssymb}
\usepackage{array,longtable}
\usepackage{listings}
\usepackage{xcolor}
\usepackage[hidelinks]{hyperref}
\usepackage{parskip}

\lstset{
  basicstyle=\ttfamily\small,
  breaklines=true,
  columns=fullflexible,
  keepspaces=true,
  frame=single,
  framerule=0.2pt,
  xleftmargin=0.6em,
  aboveskip=0.7em,
  belowskip=0.7em,
  literate={—}{{---}}1 {°}{{\ensuremath{^\circ}}}1 {≈}{{\ensuremath{\approx}}}1
           {←}{{<-}}2 {→}{{->}}2 {★}{{*}}1 {·}{{\ensuremath{\cdot}}}1
           {├}{{|--}}3 {└}{{`--}}3 {│}{{|}}1 {─}{{-}}1 {×}{{x}}1
           {Δ}{{\ensuremath{\Delta}}}1 {μ}{{\ensuremath{\mu}}}1 {ω}{{\ensuremath{\omega}}}1
           {₁}{{1}}1 {₂}{{2}}1 {²}{{\ensuremath{^2}}}1 {−}{{-}}1 {≤}{{<=}}2
           {n̂}{{\ensuremath{\hat n}}}1 {′}{{'}}1 {ᵀ}{{\ensuremath{^{T}}}}1
           {⁻¹}{{\ensuremath{^{-1}}}}2 {ᵢ}{{i}}1 {ⱼ}{{j}}1 {ε}{{\ensuremath{\varepsilon}}}1
}

\newcommand{\kw}[1]{\texttt{\uppercase{#1}}}
\newcommand{\code}[1]{\texttt{#1}}

\title{Collision Detection Science\\[2pt]
       \large Research Reference, Pure-Rust Porting Evaluation \& the
       \code{posim} Implementation}
\author{physical\_object\_simulator}
\date{July 24, 2026}

\begin{document}
\maketitle

\noindent\emph{This is the typeset twin of \code{collision\_detection.md} ---
companion to \code{grammar.pdf}, \code{physical\_object\_simulator.pdf} and
\code{scene\_info.pdf}. Written for a reader with zero prior knowledge.}

\tableofcontents

\section{What is rigid-body collision detection?}

Two solid bodies flying through space must not pass through each other. A
simulator therefore answers three questions, every time step:

\begin{enumerate}
\item \textbf{Did anything touch?} (\emph{detection}) --- and, harder: did
      anything touch \emph{during} the step, even if the step's endpoints look
      separated?
\item \textbf{Where and in which direction?} (\emph{contact geometry}) --- the
      \textbf{contact point} and the \textbf{contact normal} $\hat n$, the unit
      vector along which Newton's third law acts: body $j$ receives the impulse
      $+J\hat n$ and body $i$ receives exactly $-J\hat n$. Get the normal wrong
      and momentum leaks; get it right and the force pair is a true
      action--reaction pair.
\item \textbf{What happens next?} (\emph{response}) --- an impulse
      (instantaneous momentum exchange), a constraint force, or a stiff penalty
      spring.
\end{enumerate}

The hard parts, discovered independently by every engine surveyed:
\textbf{tunneling} (a fast body crosses a thin one between two checks),
\textbf{resting contact} (an infinite sequence of ever-smaller bounces --- the
Zeno problem), \textbf{simultaneity} (three balls touching at once --- Newton's
cradle), and \textbf{penetration recovery} (bodies that already overlap).

\section{The research: seven simulators investigated}

\subsection{Bullet / PyBullet (C++, the robotics standard)}
Broad phase: \code{btDbvtBroadphase} --- \emph{two} dynamic AABB trees (moving
vs.\ static/sleeping); alternatives \code{btAxisSweep3} (sweep-and-prune) and
brute force. Narrow phase: GJK (\code{btGjkPairDetector}) + \textbf{EPA}
penetration depth, with analytic special cases for sphere-sphere, sphere-box,
box-box (SAT). Manifold: \code{btPersistentManifold}, $\le 4$ cached points,
persisted across frames (\textbf{warm starting}), pruned by a breaking
threshold ($\approx 0.02$). Normal convention: \code{contactNormalOnB}, from B
toward A; \code{contactDistance} $< 0$ = penetration. CCD: swept-sphere
conservative advancement + \textbf{motion clamping}. Response:
\code{btSequentialImpulseConstraintSolver} (projected Gauss--Seidel), defaults
10 iterations, erp 0.2, \textbf{split impulse} below $-0.04$, restitution
suppressed under a resting-velocity threshold, warm-start 0.85, margin 0.04.
Islands, sleeping, SIMD.

\subsection{Rapier / parry (pure Rust, the modern reference)}
Broad phase: BVH over collider AABBs (quantized SIMD-friendly \code{Qbvh}).
Narrow phase: parry GJK + penetration, SAT primitives; \code{ContactManifold}s
carry per-shape local normals + a world normal, per-point \code{dist} $< 0$ =
penetration, persisted solver impulses. Normal: from collider 1 toward
collider 2. CCD: \textbf{nonlinear time-of-impact} (translation and rotation)
+ motion clamping. Response: sequential impulse with stabilization iterations,
SoA/SIMD batches, restitution threshold, islands + sleeping.

\subsection{Project Chrono / PyChrono (C++, multi-physics engineering)}
Collision systems: embedded Bullet, or the Multicore system (spatial binning +
analytic/MPR). Physics formulations: \textbf{NSC} (non-smooth --- contacts as
hard constraints, a Differential Variational Inequality solved as a Cone
Complementarity Problem by \textbf{APGD}) vs.\ \textbf{SMC} (penalty/compliant,
Hertz-style). \textbf{Envelope \& margin}: contacts anticipated at an outward
envelope before touching. Normal: the X column of the per-contact
\code{plane\_coord} rotation matrix. No dedicated CCD.

\subsection{SOFA (C++, interactive multi-physics)}
Pipeline: \code{BruteForceBroadPhase} $\to$ \code{BVHNarrowPhase} $\to$
pluggable intersection methods with \textbf{alarmDistance} (contact created
\emph{before} overlap --- the anti-tunneling band) and \textbf{contactDistance}
(the separation the response holds). \code{DetectionOutput} carries both
points, a normal (outward from the first model) and a signed distance
($< 0$ = interpenetration). Response: penalty springs or Lagrangian
unilateral constraints (Gauss--Seidel).

\subsection{Open Source Physics / EJS (Java, the teaching classic)}
Analytic 2-body overlap tests (center distance vs.\ sum of radii); the response
in closed form --- velocity decomposed along the \textbf{line of centers} (the
normal), along-normal components transformed by the 1-D elastic formulas.
Teaching devices: impact-parameter slider, the center-of-mass cross sailing
straight through the collision.

\subsection{SIMple Physics (Rust; engine = nphysics/ncollide)}
ncollide (ancestor of parry): DBVT broad phase with loosened AABBs, GJK + EPA,
\textbf{persistent manifolds} with feature IDs, proximity events. SIMple adds
a user-editable Lua \code{collision\_fn} --- students see and modify the
response formula (line-of-centers impulse; a \code{cos\_angle} $< 0$ guard
against re-colliding separating pairs).

\subsection{PhysicsHub (JavaScript, student web sims)}
Squared-distance reject, exact circle overlap, Wikipedia two-body elastic
formula along the center line, a damping slider as restitution, live KE plot.
Fixed-timestep overlap testing --- the honest example of \emph{how tunneling
happens} when nothing guards against it.

\section{The comparison chart}

\emph{(Split across two tables; same rows.)}

\begin{center}\small
\begin{tabular}{lp{3.0cm}p{3.0cm}p{3.0cm}p{2.9cm}}
 & \textbf{Bullet} & \textbf{Rapier/parry} & \textbf{Chrono} & \textbf{SOFA}\\\hline
Broad phase & dual DBVT / SAP & Qbvh BVH & DBVT or grid binning & brute-force AABB\\
Narrow phase & GJK+EPA, SAT prims & GJK+pen., SAT prims & GJK/EPA or analytic+MPR & BVH + proximity\\
Manifold & persistent $\le$4 pts & persistent + impulses & Bullet's / own & DetectionOutput list\\
Normal & on B, B$\to$A & collider1$\to$collider2 & X col of plane\_coord & outward from 1st\\
Penetration & dist $<$ 0 & dist $<$ 0 & dist $<$ 0 & dist $<$ 0\\
TOI / anti-tunnel & swept-sphere + clamp & nonlinear TOI + clamp & envelope + small dt & alarmDistance band\\
Response & seq.\ impulse PGS & seq.\ impulse & NSC (APGD) or SMC & penalty or Lagrangian\\
Stabilization & split impulse, erp 0.2 & bias iterations & constraint/compliance & contactDistance\\
Restitution & resting-vel threshold & velocity threshold & material pairs & material\\
Key numbers & margin .04, break .02 & substeps, margins & envelope/margin & alarm/contact dist\\
\end{tabular}
\end{center}

\begin{center}\small
\begin{tabular}{lp{3.6cm}p{3.6cm}p{3.6cm}}
 & \textbf{OSP / EJS} & \textbf{ncollide (SIMple)} & \textbf{PhysicsHub}\\\hline
Broad phase & none (2-body) & DBVT (loose AABBs) & squared-distance reject\\
Narrow phase & analytic circles & GJK+EPA & analytic circles\\
Manifold & single implicit & persistent + feature IDs & none\\
Normal & line of centers & GJK/EPA geometric & line of centers\\
Penetration & overlap bool & depth $\ge$ 0 & overlap bool\\
TOI / anti-tunnel & exact 2-body algebra & swept TOI & none (tunnels!)\\
Response & closed-form elastic & impulse/constraint & closed-form + damping\\
Stabilization & exact & slop + projection & none\\
Restitution & $e$ in formula & material & damping slider\\
\end{tabular}
\end{center}

\subsection{The concept tree (technique $\to$ source $\to$ posim decision)}

\begin{lstlisting}
collision detection science
|-- DETECTION
|   |-- broad phase (find candidate pairs cheaply)
|   |   |-- brute-force AABB cull        <- SOFA, PhysicsHub  -> ADOPTED (O(n^2))
|   |   |-- sweep-and-prune              <- Bullet            -> upgrade path
|   |   `-- dynamic BVH / Qbvh / binning <- Bullet/Rapier/Chrono -> upgrade path
|   |-- narrow phase (exact pair geometry)
|   |   |-- analytic primitive pairs     <- OSP lineage       -> ADOPTED
|   |   |-- SAT for boxes (15 axes)      <- Bullet box-box    -> ADOPTED
|   |   `-- GJK + EPA (any convex)       <- Bullet, parry     -> NOT needed (Sec. 4)
|   |-- contact geometry
|   |   |-- contact point                <- all               -> ADOPTED
|   |   |-- UNIT NORMAL (action-reaction)<- all               -> ADOPTED: from i to j
|   |   `-- signed separation / depth    <- all (dist < 0)    -> ADOPTED; ALSO the
|   |                                                            sundials root function
|   `-- time of impact (anti-tunneling)  * THE HARD PROBLEM *
|       |-- swept shapes + clamping      <- Bullet, Rapier    -> surveyed, not needed
|       |-- proximity band               <- SOFA, Chrono      -> surveyed
|       `-- INTEGRATOR EVENT ROOTFINDING <- (none of the 7!)  -> ADOPTED:
|                                           CVodeRootInit lands on the exact root
|-- RESPONSE
|   |-- impulse (Gauss-Seidel passes)    <- Bullet, Rapier    -> ADOPTED (K-matrix)
|   |-- restitution e + velocity thresh. <- Bullet, Rapier    -> ADOPTED (e=1, min)
|   |-- positional projection + slop     <- Bullet split imp. -> ADOPTED
|   |-- Lagrangian / APGD                <- Chrono NSC, SOFA  -> overkill
|   `-- penalty springs                  <- SOFA, Chrono SMC  -> rejected (stiffness
|                                                                fights adaptive CVODE)
`-- HURDLES (and the posim answer)
    |-- tunneling            -> rootfinding + CVodeSetMaxStep cap (verified: 100 m/s
    |                           vs 5 mm plate)
    |-- resting contact/Zeno -> tiered guard: 64 events -> plastic; 128 -> disarm+project
    |-- simultaneity         -> Gauss-Seidel over all pairs (Newton's cradle emerges)
    |-- deep initial overlap -> end-of-interval sweep + projection
    `-- degenerate geometry  -> skipped safely, never a division by zero
\end{lstlisting}

\section{Evaluation for a pure-Rust port (zero unsafe, zero dependencies)}

\textbf{Essential --- adopted:} AABB broad-phase cull; analytic narrow phase
(with a \emph{closed} shape set \{point, sphere, cuboid\} every pair has an
exact closed form --- faster than GJK, no termination heuristics, and the
normal is \textbf{exact}, which is the entire point); SAT for cuboid-cuboid;
the impulse response with the effective-mass matrix $K$ (provably
momentum-conserving action--reaction along $\hat n$, angular terms give correct
spin transfer); restitution with a velocity threshold; positional projection
with slop; sequential Gauss--Seidel multi-pair passes; and --- \emph{novel
here} --- \textbf{integrator event rootfinding for the time of impact}: the
integration backend (SUNDIALS) already contains a root finder that checks every
internal step and interpolates onto the crossing, a precision no engine-side
swept test can match.

\textbf{Valuable --- documented upgrade tier:} persistent manifolds with warm
starting (stacking); 2--4-point face manifolds (boxes resting flat);
sweep-and-prune/BVH broad phase (hundreds of bodies); Coulomb friction
(per-object $\mu$, tangential impulse clamped to the cone --- designed,
deliberately not shipped in this delivery); islands + sleeping.

\textbf{Overkill --- rejected with reasons:} GJK/EPA/MPR (no open shape set to
serve); NSC complementarity solvers (APGD --- research-grade machinery for
granular matter); penalty springs (their stiffness would fight CVODE's adaptive
error control); GPU/SIMD batching (no workload to feed it).

\emph{Why no GJK/EPA} (the one decision an expert would question): GJK exists
to handle \emph{arbitrary} convex shapes behind a support function. posim's
shape set is closed, so all six pair types have exact constant-time formulas
with exact normals; Bullet itself special-cases exactly these pairs. If a
general convex hull is ever added, GJK+EPA (or parry) becomes the right tool;
the \code{Contact} record and the solver do not change. \emph{(The shape set
has since grown --- torus, disk, cylinder --- and remains closed:
Section~6's three-tier dispatch keeps every ball contact and every face
contact exact, still without GJK.)}

\section{The recommendation: collision detection in ALL rigid-body scenes}

\begin{enumerate}
\item \textbf{On by default, everywhere.} Every \kw{step}, \kw{run} and
      scene-window playback detects collisions whenever a collidable pair
      exists; \kw{collide off} opts out. Two \code{Point}s cannot collide; two
      static bodies are skipped.
\item \textbf{Event-driven time of impact.} The pairwise \textbf{signed
      separation} is the integrator's root function (\code{CVodeRootInit} /
      \code{ARKodeRootInit}). CVODE/ARKODE check it after every internal step
      and interpolate onto the zero crossing: measured impact-time errors are
      at the $10^{-15}$ relative level. Approach-only direction filtering plus
      the integrator's zero-at-restart handling prevent re-triggering.
\item \textbf{Anti-tunneling cap.} While armed, the internal step is capped
      at (smallest feature)/(2 $\times$ max \emph{reachable} relative
      \emph{surface} speed) --- center speed plus what the accelerations
      can add before the next refresh, plus each body's spin bound times
      $R_{\mathrm{bound}}$, since a rotating edge can sweep into contact
      without the centers moving --- refreshed after every event and at
      every interval start (Section~6 has the exact bounds).
\item \textbf{Exact impulse at the root.}
      $J = -(1+e)\,(\mathbf v_{\mathrm{rel}}\!\cdot\!\hat n)\,/\,
      (\hat n^{\!\top} K \hat n)$ with
      $K = (m_i^{-1}\!+m_j^{-1})\,\mathbf 1
      - [r_i]_\times I^{-1}_{i,w} [r_i]_\times
      - [r_j]_\times I^{-1}_{j,w} [r_j]_\times$,
      applied as $\pm J\hat n$ (and $\pm r \times J\hat n$) through the
      canonical setters --- the action--reaction pair by construction. Static
      bodies receive no writes. $e = \min(e_i, e_j)$, default 1.
\item \textbf{Guards.} Tiered Zeno guard (64 events $\to$ plastic; 128 $\to$
      disarm + project); \emph{elastic} end-of-interval penetration sweep
      (deep initial overlaps and grinding approximate contacts); degenerate
      geometry skipped safely.
\item \textbf{The normal is public.} Grammar paths (\code{contact0.normal}),
      \kw{contacts}, machine-mode JSON, and golden arrows in the scene window.
\item \textbf{Structural safety.} With zero collidable pairs the rootfinding is
      never armed and the CVODE code path is \textbf{bit-identical} to the
      pre-collision implementation (verified in the sundials source and
      enforced by test).
\end{enumerate}

\section{The posim implementation}

\textbf{Modules.} \code{physical\_object/src/boundary.rs} (the shape set:
SDFs, support extents/points/ranks, analytic inertia);
\code{physical\_object/src/collide.rs} (geometry + impulse response, fully
unit-tested); event wiring in \code{physical\_object/src/integrate.rs} (CVODE
and SPRK event loops); grammar in \code{posim} (\kw{collide}, \kw{contacts},
\kw{box}, \code{contactK} paths); scene arrows + shape wireframes in
\code{posim/src/scene/}.

\textbf{The shape set.} Six boundaries: \code{Point}, \code{Sphere},
\code{Cuboid}, and --- since this release --- \code{Torus} (a centerline
circle of radius $c$ in the local $xy$-plane swept by a tube of radius $a$;
inner radius $c-a$, outer radius $c+a$), \code{Disk} (an \textbf{ideal
zero-thickness} solid disk in the local $xy$-plane), and \code{Cylinder}
(radius $r$, half-height $h$; full height $2h$). Each new shape carries an
\textbf{exact SDF} (torus: distance to the centerline circle minus the tube
radius; disk: the unsigned distance, zero exactly on the disk --- a
zero-thickness body has no interior; cylinder: the 2-D box SDF in
$(\rho, z)$) and the \textbf{analytic inertia tensor}: torus
$I_z = m(c^2 + \tfrac34 a^2)$, $I_x = I_y = m(\tfrac12 c^2 + \tfrac58 a^2)$;
disk $I_x = I_y = \tfrac14 m a^2$, $I_z = \tfrac12 m a^2$
(perpendicular-axis theorem); cylinder $I_z = \tfrac12 m r^2$,
$I_x = I_y = m(3r^2 + 4h^2)/12$. Every shape also provides its
\textbf{support extent} $h(u)$ in exact closed form (for the torus, the
support of its convex hull --- a support function cannot see the hole), a
\textbf{support point} achieving it --- the \textbf{centroid of the
supporting set} when that set is a face, edge, circle or cap, so a flat-on
contact carries no spurious lever arm --- and a \textbf{support rank}
(0 = single point/rim point, 1 = edge or circle, 2 = face or cap).

\textbf{The Contact record} (pinned convention, ARCHITECTURE.md \S 3.8):

\begin{lstlisting}
pub struct Contact {
    pub i: usize,       // first body (lower index)
    pub j: usize,       // second body
    pub t: f64,         // event time (the interpolated root)
    pub point: Vec3,    // world-space contact point
    pub normal: Vec3,   // UNIT normal, from body i toward body j
    pub depth: f64,     // penetration depth >= 0 (~ 0 at a root)
    pub rel_vel_n: f64, // pre-impulse (v_j - v_i).n  (< 0 = approaching)
    pub impulse_n: f64, // scalar impulse magnitude J
}
\end{lstlisting}

\textbf{The three-tier dispatch} (still no GJK/EPA/MPR machinery):
\begin{enumerate}
\item \textbf{Exact ball tier} --- a \emph{ball} (sphere, or point as the
      zero-radius case) against \textbf{anything}: separation = the partner
      shape's exact SDF at the ball center (in the shape frame) minus the
      ball radius. Sphere-sphere keeps $|\Delta| - (r_a + r_b)$ with the
      normal along the line of centers; sphere-cuboid keeps the box-SDF
      formula (continuous inside and out; the interior branch picks the
      nearest face). For torus/disk/cylinder the contact is the exact
      closest surface point, the normal the exact outward direction, and
      the contact point sits midway across the overlap band. Because the
      SDFs are exact, a small ball \textbf{can genuinely thread a torus
      hole} (its separation never reaches zero on a through-hole line) and
      a point particle \textbf{passes through the ideal zero-thickness
      disk} (a measure-zero contact --- see the FAQ). Degenerate
      configurations (ball center on the torus centerline, exactly on the
      disk, on the cylinder axis at side contact) are skipped safely.
\item \textbf{Exact cuboid SAT} --- cuboid-cuboid is the 15-axis SAT,
      \textbf{unchanged}: the maximum separation is the signed separation,
      its axis (oriented $i \to j$) the normal; contact point = deepest
      support vertex (face case) or the closest points of the supporting
      edges (edge case).
\item \textbf{Support-axis tier} --- every remaining extended-vs-extended
      pair (any pair involving a torus/disk/cylinder not covered above).
      The SAT gap $|d \cdot l| - (h_a(l) + h_b(l))$ is evaluated over a
      finite candidate-axis list: each cuboid's three \textbf{face axes},
      each round shape's \textbf{symmetry axis}, every normalized
      \textbf{cross product} of those primary axes (edge-vs-edge and
      edge-vs-rim separations; near-parallel pairs skipped as in the SAT),
      the \textbf{radial rejection axes} --- the component of the center
      offset perpendicular to each primary axis, which is the true lateral
      separating direction when two round shapes lie (near-)parallel with
      an axial offset, exactly where the parallel-axis cross product
      vanishes and the raw center line is tilted --- and the \textbf{center
      line}; the maximum gap is the separation, its axis (oriented
      $i \to j$) the normal. \textbf{Exact for face-on contacts} --- in
      particular every wall-slab contact in a \kw{box} --- and
      \textbf{exact for side-side contacts of parallel and near-parallel
      round shapes}: two parallel cylinders with overlapping axial ranges
      separate by exactly the lateral gap, with a purely lateral normal
      (test \code{parallel\_cylinders\_side\_contact\_is\_exact}; this used
      to fire early with a tilted normal). The conservative caveat now
      applies \textbf{only to genuinely skew corner-on configurations}
      (contact may be reported slightly early, along a candidate axis);
      the torus is seen as its convex hull, so only balls can thread the
      hole. One face-on configuration stays invisible outright: two disks
      with \textbf{parallel} planes both have zero extent along the shared
      normal, so their separation is $|dz|$ --- it touches zero at plane
      coincidence without a sign change, exactly like the
      point-through-disk case, and the crossing produces no root for the
      downward-crossing rootfinder (pinned by test
      \code{parallel\_disk\_disk\_separation\_is\_the\_documented\_limitation};
      tilt one disk or model a thin cylinder --- see the FAQ).
      \textbf{Contact point = the support point of the lower-support-rank
      body} --- the incident body's deepest point (a tilted cylinder's rim
      point against a wall face, not the centroid of the wall face).
      \textbf{Equal ranks prefer the body with the clearly smaller flat
      footprint} (\code{support\_footprint\_radius}, the lateral
      circumradius of the supporting set: chosen when under half the
      partner's), so a small cap landing on a big wall face contacts at
      the \textbf{cap center}, not halfway toward the face centroid (test
      \code{small\_cap\_on\_large\_face\_contacts\_at\_the\_cap\_center});
      comparable footprints use the midpoint of the two support-set
      centroids.
\end{enumerate}

\textbf{The rigid box (\kw{box} \code{<size>}).} Six static \code{Cuboid}
wall slabs enclosing an axis-aligned cube of the given inner side length,
each with \code{inverse\_mass} $= 0$ \textbf{and} zero inverse inertia. This
is infinite mass \emph{exactly as the equations of motion see it}: the state
stores momenta and the dynamics only ever multiply by the inverse ---
$v = p\,m^{-1}$, and the impulse denominator is
$\hat n^{\!\top} K \hat n = m_i^{-1} + m_j^{-1} + \text{angular terms}$. A
static side therefore contributes exactly 0 to every denominator, receives
no state writes at all, and reflects everything elastically while remaining
bit-identically at rest --- no large-mass approximation appears anywhere.

\textbf{The root function \emph{is} the narrow phase}: the same
\code{separation\_at} geometry runs inside the CVODE/ARKODE $g$ function on the
packed state, so detection and response can never disagree.

\textbf{Anti-tunneling with spin --- and with acceleration.} The
while-armed step cap is (smallest crossable feature)/(2 $\times$ max
\emph{reachable} relative \emph{surface} speed), where \emph{reachable}
spans the horizon $\Delta t$ to the next refresh --- one output interval
(the cap is re-refreshed at every interval start and after every event).
The linear term is the relative center speed \textbf{plus
$(a_i + a_j)\,\Delta t$}, each body's acceleration taken from pairwise
gravity at the current configuration, uniform gravity, $qE$ and the
external force (the magnetic force is $\perp v$ and can never grow speed)
--- so a ball released \textbf{from rest} above a thin plate, whose
instantaneous relative speed is zero at arm time, still gets a finite cap
and is caught (test
\code{ball\_released\_from\_rest\_does\_not\_tunnel\_the\_thin\_plate}:
TOI matches the analytic $\sqrt{2h/g}$ fall to $10^{-6}$). The spin term
bounds each body's rate as
$|\omega| \le \sqrt3\,\|I^{-1}\|_\infty\,\bigl(|L| + \Delta t\,(|\tau_{\mathrm{ext}}|
+ \|M\|_\infty |B|)\bigr)$, multiplied by its $R_{\mathrm{bound}}$ --- a
bound through the angular \emph{momentum} rather than the instantaneous
$|\omega|$, so it covers torque-free polhode motion exactly ($L$ is
conserved, yet $|\omega|$ can spike up to $|L|/I_{\min}$ mid-interval with
no torque at all). The cap returns no-cap only when literally nothing can
move. Feature sizes of the new shapes: the torus's tube radius, the
cylinder's $\min(r, h)$; the ideal disk has zero thickness of its own, so
its cap comes from the radius of whatever ball approaches it (the pairwise
min picks the smaller feature).

\textbf{The elastic closing sweep.} \code{resolve\_penetrations} now takes an
explicit mode: the Zeno tier-2 guard still resolves \textbf{plastically}
(kill a chattering contact dead), but the end-of-interval safety sweep
resolves \textbf{elastically} --- a slowly-grinding approximate
(support-axis) contact that never produces a clean root can no longer bleed
energy sweep after sweep.

\textbf{The hmax-span fix (an honest bug report).} The after-event refresh of
the anti-tunneling cap clamps the computed \code{hmax} into a span, and that
span used to be $t_{\mathrm{out}} - t$ --- the remainder of the current
\emph{output interval}. A collision landing exactly \textbf{on} an output
boundary (which the box geometry of Example 11 promptly produced) collapsed
that span to zero, pinning the max step at the $10^{-12}$ clamp floor; the
stale cap was never recomputed, so every later interval was starved into
\code{CV\_TOO\_MUCH\_WORK} (500\,000 no-advance steps). The fix is twofold:
the clamp span is now the remaining \textbf{run} ($t_{\mathrm{end}} - t$,
exactly as at arm time), and the cap is re-refreshed at the start of every
output interval, so a cap computed near a previous $t_{\mathrm{out}}$ can
never go stale.

\textbf{Defaults.} restitution 1.0 (elastic), combine $\min(e_i,e_j)$;
\code{restitution\_threshold} $10^{-3}$; \code{contact\_slop} $10^{-9}$;
Zeno tiers 64 / 128 events per output interval.

\section{Command reference}

\begin{center}
\begin{longtable}{p{5.6cm}p{8.6cm}}
\textbf{command / path} & \textbf{meaning}\\\hline
\kw{collide} & report status: on/off, collidable pairs, impulses so far\\
\kw{collide on} / \kw{collide off} & enable (default) / disable detection\\
\kw{contacts} & list every contact of the last \kw{step}/\kw{run}\\
\kw{new torus} \code{\{ \ldots \}} & new torus: \code{ring\_radius}/\code{tube\_radius} (defaults 1 / 0.25) \textbf{or} the \code{inner\_radius} + \code{outer\_radius} pair --- resolved and validated once at the closing brace, so the pair is genuinely order-independent; \code{inner\_radius} $= 0$ (the horn torus) is valid\\
\kw{new disk} \code{\{ \ldots \}} & new \textbf{ideal zero-thickness} solid disk: \code{radius} (default 1)\\
\kw{new cylinder} \code{\{ \ldots \}} & new cylinder: \code{radius} (default 0.5); \code{height} sets the \emph{full} height (default half\_height 1)\\
\kw{new cube} / \kw{new disc} & aliases for \kw{cuboid} / \kw{disk}\\
\kw{box} \code{<size>} / \kw{box off} / \kw{box} & create / remove / report the rigid bounding box: six static wall slabs with \code{inverse\_mass} $= 0$ (infinitely massive)\\
\code{objN.ring\_radius}, \code{.tube\_radius} & the torus generating radii (writes recompute inertia)\\
\code{objN.inner\_radius}, \code{.outer\_radius} & the derived torus pair (each write holds the other one fixed)\\
\code{objN.height} / \code{.half\_height} & cylinder height (\code{height} $= 2 \times$ half\_height)\\
\code{objN.radius} & now serves sphere, \textbf{disk and cylinder} (a write keeps the shape family); a torus write errors --- set \code{ring\_radius}/\code{tube\_radius} or \code{inner\_radius}/\code{outer\_radius} instead\\
\code{objN.inverse\_mass} & 0 for a wall slab --- infinite mass, exactly\\
\code{GET system.box} & inner box size (0 = none)\\
\code{GET contactK.i} / \code{.j} & the colliding pair (indices)\\
\code{GET contactK.t} & event time (the interpolated root)\\
\code{GET contactK.point} & world contact point (vec3)\\
\code{GET contactK.normal} & \textbf{unit action--reaction line, $i \to j$}\\
\code{GET contactK.depth} & penetration depth at the event\\
\code{GET contactK.rel\_vel\_n} & approach speed along the normal (negative)\\
\code{GET contactK.impulse} & scalar impulse magnitude $J$\\
\code{SET objN.restitution = e} & bounciness 0 (plastic) \ldots{} 1 (elastic, default)\\
\code{GET system.contacts} & number of recorded contacts\\
\code{GET system.collisions} & running impulse total this session\\
\code{SET system.restitution\_threshold} & resting-contact speed threshold\\
\code{SET system.contact\_slop} & tolerated overlap before projection\\
\end{longtable}
\end{center}

Contact paths are ordinary expression atoms:
\code{contact0.impulse * contact0.normal.x} computes the x-component of the
impulse vector. Machine mode's \code{\{"op":"state"\}} includes
\code{"contacts"}, \code{"collide\_enabled"}, \code{"collision\_count"}, and
--- since this release --- \code{"box"} (number or null) plus per-object
\code{"wall"} and \code{"inverse\_mass"}. The scene window's
\textbf{Contacts} button / \code{C} key toggles golden normal arrows at the
contact points.

\kw{list} tags each wall slab \code{[wall: static, inverse\_mass=0]} (any
other inverse-mass-0 body reads \code{[static, inverse\_mass=0]}); \kw{del}
keeps the wall index list renumbered, and deleting any wall dissolves the
box (\code{system.box} $\to$ 0) while the surviving slabs \textbf{stay
tracked}: \kw{list} keeps their \code{[wall]} tag, bare \kw{box} reports
\code{box: dissolved (a wall was deleted; N tracked slab(s) remain ---
BOX <size> replaces them, BOX OFF removes them)}, and \kw{box}
\code{<size>} removes the survivors before building the new box --- no
orphan slab is ever leaked. Scene \code{init} entities carry the
new shape parameters and \code{"wall": true}, with a top-level
\code{"box": <size>}; \code{scene.html} draws torus (outer/inner equators +
tube rings + 4 cross-sections), disk (rim + 2 diameters) and cylinder
(2 rims + 4 side lines) wireframes, all quaternion-rotated so spin is
visible, plus a dashed \code{\#5d84a8} interior wireframe for the box ---
the wall slabs themselves are \textbf{not} drawn as bodies.

\section{How it was verified}

37 unit tests (geometry: known separations, SAT face vs.\ edge, continuity
across touch; response: $\hat n \!\cdot\! K \hat n = m_1^{-1}+m_2^{-1}$ for
central hits, static-wall reflection with the wall bit-unchanged, separating
pairs untouched, projection separating deep overlap; new this release: the
torus/disk/cylinder \textbf{SDFs at hand-checked points}, support
extents/points (the invariant $p \cdot u = h(u)$ verified across shapes and
directions) and ranks, the new \textbf{analytic inertia tensors} (torus
$I_z = 2.4375$ for $m = 1$, $c = 1.5$, $a = 0.5$; disk $I_z = I_x + I_y$),
ball-vs-torus contact with the hole genuinely passable, disk face/rim and
cylinder side/cap contacts, the \textbf{tilted-torus anchor} --- a flat
torus of outer radius 2 exactly inscribes a 4-wide box (separation 0 to
$10^{-12}$) while the $(1,1,1)/\sqrt3$ tilt clears every wall by exactly
$2 - (1.5\sqrt{2/3} + 0.5) \approx 0.2753$ --- and a static slab
\textbf{reflecting a cylinder elastically} with the wall bit-unchanged;
new in the review round: \textbf{exact side-side contacts of parallel
cylinders} via the radial rejection axes
(\code{parallel\_cylinders\_side\_contact\_is\_exact} --- the lateral gap
to $10^{-12}$ across four offsets and a purely lateral normal), the
\textbf{pinned parallel disk-disk limitation}
(\code{parallel\_disk\_disk\_separation\_is\_the\_documented\_limitation}
--- separation $= |dz|$ exactly, touching zero without a sign change) and
the \textbf{smaller-footprint contact rule}
(\code{small\_cap\_on\_large\_face\_contacts\_at\_the\_cap\_center} --- a
0.25-radius cap pressed into a big slab face contacts at the cap center)).
15 integration tests, each against a closed form: velocity exchange with
\textbf{TOI error} $< 10^{-9}$, unequal-mass formulas, restitution ratios and
plastic KE loss $\tfrac12 \mu v^2$, oblique normals = line of centers,
off-center $\Delta L = r \times J$ with total $L$ conserved, 3-sphere cradle,
bounce apex $= e^2 h$ at $t = \sqrt{2h/g}$, \textbf{no tunneling} at 100 m/s
vs.\ a 5 mm plate, SPRK-path exchange, \textbf{bit-identical zero-pair
invariance}, armed-but-quiet agreement $\le 10^{-12}$; new this release: a
\textbf{ball in a rigid box} of inverse-mass-0 slabs (energy conserved to
$10^{-9}$, the ball never escapes, all six walls bit-identical to
construction), a \textbf{point particle threading the torus hole} with zero
events while a fat ball on a nearby line hits the tube at the analytic
$\mathrm{TOI} = (3 - \sqrt{1.45})/4$ and the free torus recoils ($E$ and $p$
conserved), and a \textbf{mixed cylinder + disk + cube rattle} inside the
box with $|\mathrm dE|/E < 10^{-6}$; new in the review round: a ball
\textbf{released from rest} above a thin static plate under uniform
gravity, caught at the analytic free-fall
$\mathrm{TOI} = \sqrt{2 \cdot 4.985/10}$ to $10^{-6}$
(\code{ball\_released\_from\_rest\_does\_not\_tunnel\_the\_thin\_plate}
--- the acceleration-aware anti-tunneling cap at work: the instantaneous
relative speed is zero at arm time). Grammar tests (33 posim tests) cover
the whole command family including error paths, now with the
\kw{new torus}/\kw{disk}/\kw{cylinder} parameter paths (the
order-independent inner/outer pair, \kw{height} as the full height,
\code{radius} across the shape family) and the whole \kw{box} family
(creation, status, \kw{off}, \code{system.box}, wall tags, machine-mode
\code{box}/\code{wall}/\code{inverse\_mass}); new in the review round:
\code{torus\_pair\_is\_order\_independent\_and\_new\_is\_transactional}
(both orders of the inner/outer pair --- including
\code{\{ outer\_radius = 0.5, inner\_radius = 0.2 \}}, which used to fail
in one order --- the horn torus \code{inner\_radius} $= 0$ on \kw{new} and
\kw{set}, the refused \code{SET objN.radius} on a torus, and a failed
\kw{new} leaving \textbf{no ghost object} behind) and
\code{box\_recreate\_after\_wall\_deletion\_leaks\_nothing} (bare \kw{box}
reports the dissolved state; \kw{box} \code{<size>} after a wall deletion
removes the surviving tracked slabs first). The scene frame was verified
live to carry the exact analytic normal/point/impulse, and the arrow's
pixels were verified on the canvas; for this release the box-of-shapes demo
was re-verified in a real browser: the \code{init} frame carried
\code{box: 4}, the wall flags and every new shape's parameters; the window's
playback copy conserved $E$ to $10^{-9}$; the golden contact arrows drew
(2218 gold pixels counted on the canvas) and the dashed box wireframe drew
(5904 pixels); reverse playback was exercised. All regression anchors (outer
solar system, Kepler, gyroradius, tumbling body) still print SUCCESS;
\textbf{94 tests green} (37 lib + 15 collision + 9 conservation +
33 posim); the workspace builds with zero warnings; \code{Cargo.lock} still
lists only the five local crates.

\section{The eleven examples (run for the USER and the MANAGER)}

All scripts live in \code{scripts/collisions/} and run with\\
\code{cargo run -p posim -- --script scripts/collisions/NN\_name.posim}.\\
The outputs below are the real captured runs. Two additional self-checking Rust
examples (\code{newtons\_cradle}, \code{bouncing\_ball\_restitution}) print
SUCCESS/FAILURE and exit nonzero on failure.

\subsection{Example 1 --- head-on exchange (the canonical action--reaction pair)}
Two identical spheres approach at $\pm 1$; gap 3, closing speed 2 $\Rightarrow$
impact at $t = 1.5$ exactly; velocities exchange, momentum stays 0, energy is
conserved.
\begin{lstlisting}
Out[7]= t = 3 (advanced by 3, 26 solver steps, 1 collision(s) - CONTACTS lists them)
Out[8]= contact0: obj0 <-> obj1 at t = 1.5
  point  = [0, 0, 0]
  normal = [1, 0, 0]  (from obj0 toward obj1)
  depth = 0, approach speed = 2, impulse = 2
get obj0.velocity -> [-1, 0, 0]     get obj1.velocity -> [1, 0, 0]
energy -> 1 (unchanged)
\end{lstlisting}

\subsection{Example 2 --- unequal masses (the light ball bounces back)}
$m_1 = 1$ at $v = 1$ strikes $m_2 = 3$ at rest:
$v_1' = (m_1{-}m_2)/(m_1{+}m_2) = -0.5$,
$v_2' = 2m_1/(m_1{+}m_2) = +0.5$.
\begin{lstlisting}
get obj0.velocity.x -> -0.5      get obj1.velocity.x -> 0.5
momentum -> [1, 0, 0] (conserved)
\end{lstlisting}

\subsection{Example 3 --- the restitution ladder}
The same collision with $e \in \{1.0, 0.8, 0.5, 0.2\}$: separation speed
$= e \times 2$ and kinetic energy $= e^2$ in every rung.
\begin{lstlisting}
e = 1.0:  separation 2      energy 1
e = 0.8:  separation 1.6    energy 0.64
e = 0.5:  separation 1      energy 0.25
e = 0.2:  separation 0.4    energy 0.04
\end{lstlisting}

\subsection{Example 4 --- Newton's cradle (five spheres)}
Four touching balls, one incomer at $v = 1$; the Gauss--Seidel passes propagate
the impulse through the simultaneous contacts:
\begin{lstlisting}
t = 4 (..., 4 collision(s))
v0..v3 -> 0, 0, 0, 0     v4 -> 1     momentum -> [1, 0, 0]
\end{lstlisting}
Only the far ball exits, at the incomer's speed.

\subsection{Example 5 --- billiard break (every outgoing direction is a normal)}
A cue ball at $v = 2$ breaks a three-ball triangle at $e = 0.95$:
\begin{lstlisting}
contact0: normal [1, 0, 0]              (cue -> apex, head on)
contact1: normal [0.835.., 0.55, 0]     (apex -> upper ball)
contact2: normal [0.835.., -0.55, 0]    (apex -> lower ball)
momentum before -> [2, 0, 0]    after -> [2, 0, 0]  (exact)
\end{lstlisting}

\subsection{Example 6 --- off-center spin-up ($L = r \times J\hat n$)}
A sphere strikes a free cuboid above its center; contact at
$[-0.5, 0.5, 0]$ with $J = 4.444\ldots$ along $+x$:
\begin{lstlisting}
get obj1.angular_momentum -> [0, 0, -2.2222222222222223]
\end{lstlisting}
$r \times J\hat n = (0, 0, -0.5 \cdot 4.444) = (0, 0, -2.222)$ \checkmark;
total angular momentum about the origin stays $-2$ through the transfer.

\subsection{Example 7 --- the thin-wall time of impact (the tunneling killer)}
A 1 cm bullet at speed 100 meets a 5 mm static plate inside a single 0.05 s
output step; analytic TOI $= (2 - 0.005 - 0.01)/100 = 0.01985$:
\begin{lstlisting}
get contact0.t      -> 0.019850000000000846   (error ~ 8e-16!)
get obj1.velocity.x -> -100                   (reflected exactly)
get obj1.position.x -> -3.03                  (back on its own side)
\end{lstlisting}
A fixed-step engine would tunnel; the integrator's root finder cannot miss.

\subsection{Example 8 --- colliding gravitational binary}
Two spheres on a bound orbit whose pericenter (0.5) is below the touching
distance (0.6), $e = 0.6$; the event fires on the \emph{curved} trajectory:
\begin{lstlisting}
contact0 at t = 2.884...  normal [-0.556, 0.831, 0]  (line of centers at TOI)
energy  -0.4 -> -0.4996   (the inelastic bounce shrinks the orbit)
momentum -> [2.8e-16, -2.2e-16, 0]  (zero to rounding)
\end{lstlisting}

\subsection{Example 9 --- the spinning target ($\omega \times r$ terms at work)}
The cuboid tumbles at $\omega_z = 2$; the struck face carries its own velocity:
\begin{lstlisting}
contact0: normal [-0.986, 0.167, 0]   (the ROTATED face normal)
angmom before -> [0, 0, 0.16]   after -> [0, 0, 0.15999999999999956]
energy before -> 2.96           after -> 2.9599999999999995
\end{lstlisting}
Angular momentum and energy conserved through a rotating-surface impact.

\subsection{Example 10 --- the billiard box (seven bounces, zero drift)}
An elastic ball ping-pongs between two static walls for 20 time units:
\begin{lstlisting}
t = 20 (356 solver steps, ..., |dE/E| = 0.000e0, 7 collision(s) ...)
get system.collisions -> 7      energy -> 1.4449999999999998 (unchanged)
get obj2.velocity.x   -> -1.7
\end{lstlisting}
Also the scene demo: \kw{scene create} + \kw{scene start} shows a golden
normal arrow at every wall strike.

\subsection{Example 11 --- the box of shapes (every shape in the infinitely
massive box)}
$R = 1$, $M = 1$. \kw{box} 4 builds the rigid box; six bodies fly inside it:
a \textbf{torus} $M$ (inner radius 1, outer radius 2), a \textbf{point} $M$
at $v = (100, 200, 100)$ --- the only mover, so
$E_0 = \tfrac12 \cdot 1 \cdot 60000 = 30000$ exactly --- a \textbf{sphere}
$2M$ of $r = 1/2$ (the spec leaves the radius free; $R/2$ is documented in
the script), a tilted \textbf{disk} $2M/3$ of $r = 1$, a \textbf{cube}
$5M/3$ of side 1, and a tilted \textbf{cylinder} $2M$ of $r = 1/2$, height
$3/2$. The geometry is sharp: an axis-aligned torus of outer radius 2
\textbf{exactly inscribes} the 4-box (its outer equator touches four walls),
so ``inside and not touching'' forces a tilt --- with the axis along the
body diagonal $(1,1,1)/\sqrt3$ the extent per axis is
$1.5\sqrt{2/3} + 0.5 \approx 1.7247 < 2$ (clearance 0.2753) and the torus
sits at the center. The other five positions are drawn by a documented LCG
($x \leftarrow 1664525\,x + 1013904223 \bmod 2^{32}$, seed 20260724,
sequential rejection at $\ge 0.05$ separation), so the script is
reproducible, not hand-tuned.
\begin{lstlisting}
In[2]:= box 4
Out[2]= box: inner size 4 x 4 x 4 — six static walls obj0, obj1, obj2, obj3, obj4, obj5 with inverse_mass = 0 (infinitely massive); objects collide elastically off the inside faces
In[10]:= collide
Out[10]= collisions ON (51 collidable pair(s); 0 impulse(s) so far)
get obj0.inverse_mass   -> 0
get obj6.inertia_tensor -> [[1.28125, 0, 0], [0, 1.28125, 0], [0, 0, 2.4375]]
energy -> 30000         momentum -> [100, 200, 100]
In[15]:= run 0.1 steps 100
Out[15]= t = 0.1 (2121 solver steps, 100 snapshots, |dE/E| = 2.040e-10, 119 collision(s) — CONTACTS lists them)
energy   -> 29999.99999388012
momentum -> [146.4891109312449, 102.14781200712048, 46.01601279520784]
\end{lstlisting}
\textbf{119 impulses across all three dispatch tiers in 0.1 time units, and
the energy drift is $|\mathrm dE/E| = 1.031 \cdot 10^{-9}$.} The 51
collidable pairs are the 66 pairs of 12 bodies minus the 15 static
wall--wall pairs; the torus inertia readback matches the analytic tensor
exactly ($m(\tfrac12 c^2 + \tfrac58 a^2) = 1.28125$,
$m(c^2 + \tfrac34 a^2) = 2.4375$ for $c = 1.5$, $a = 0.5$).
\textbf{Momentum is NOT conserved --- $(100, 200, 100) \to
(146.49, 102.15, 46.02)$ --- and is not supposed to be}: every wall impulse
transfers $\pm J\hat n$ into a body with \code{inverse\_mass} $= 0$. The
infinitely massive walls absorb momentum at zero velocity, so they absorb no
energy ($p^2/2m \to 0$ as $m \to \infty$, taken \emph{exactly} by the
inverse-mass formulation), and after 119 collisions the six walls are
\textbf{bit-identically at rest}. Momentum non-conservation with exact
energy conservation is the physical signature of a rigid container --- not a
bug.

\section{FAQ}

\textbf{Do I have to turn anything on?} No --- collisions are on by default in
every scene; \kw{collide off} disables, \kw{collide} reports.

\textbf{How precise is the impact time?} Interpolated onto the root: measured
errors at the $10^{-15}$ relative level (Examples 1 and 7).

\textbf{Three bodies touching at once?} The solver sweeps all pairs until
nothing still approaches (Example 4).

\textbf{Can objects still tunnel?} Not on the CVODE paths (step capped
against the smallest feature, and the cap counts speed the accelerations
can build before the next refresh --- even a from-rest drop is caught). On
SPRK the sampling bound is the fixed dt --- choose it below (thinnest
body)/(fastest approach).

\textbf{Why is momentum not conserved in the box?} Because the walls are
infinitely massive, and infinite mass absorbs momentum without absorbing
energy: a wall receiving impulse $J$ changes its momentum by $J$ but its
kinetic energy by $p^2/2m \to 0$ as $m \to \infty$. Ordinarily that is an
approximation; here the limit is \emph{exact}, because the equations of
motion only ever use the \textbf{inverse} mass --- $v = p\,m^{-1}$, impulse
denominator $m_i^{-1} + m_j^{-1} + \text{angular terms}$ --- and a wall with
\code{inverse\_mass} $= 0$ contributes zero to every denominator and
receives no state writes at all. Newton's third law still acts at every
contact; the ``missing'' momentum went into the box (Example 11:
$(100, 200, 100) \to (146.49, 102.15, 46.02)$ while energy is conserved to
$1.031 \cdot 10^{-9}$ and the walls stay bit-identically at rest).

\textbf{Can anything pass through the torus hole?} Balls can.
Ball-vs-anything uses the exact SDF, and on a line through the hole a small
ball's separation dips toward (inner radius $-$ ball radius) without ever
reaching zero --- no event fires. Verified by test: a point particle threads
the hole with zero collisions while a fat ball on a nearby line hits the
tube at the analytic $\mathrm{TOI} = (3 - \sqrt{1.45})/4$. Extended shapes
(cuboid, disk, cylinder, another torus) are served by the support-axis tier,
which sees the torus's \textbf{convex hull} --- a support function cannot
see a hole --- so only balls can thread it.

\textbf{Why does the point particle pass through the disk?} The disk is
\emph{ideal}: a zero-thickness solid disk whose ``SDF'' is the unsigned
distance to it --- continuous, and zero exactly on the disk. Against a
zero-radius point the separation touches zero for a single instant and rises
again without ever becoming negative: there is no sign change for the root
finder and no overlap to resolve --- a measure-zero contact carrying no
impulse, which is the exact answer for these two idealized bodies. Give
either party finite extent \textbf{along the contact normal} --- any
sphere radius, a cuboid, a cylinder --- and the separation genuinely
crosses zero and the contact fires (every finite-size body in Example 11
bounces off the disk). The one pairing with zero extent on \emph{both}
sides is \textbf{two disks with parallel planes}: their separation is
$|dz|$, which touches zero at plane coincidence without a sign change, so
a face-on disk-disk crossing is exactly as invisible as the
point-through-disk case (pinned by test
\code{parallel\_disk\_disk\_separation\_is\_the\_documented\_limitation}).
Workaround: tilt one disk (a tilted disk has finite extent along the
other's normal, so the separation genuinely crosses zero) or model a thin
cylinder.

\textbf{Friction?} Designed (per-object $\mu$, Coulomb cone clamp) but
deliberately not shipped in this delivery; the exact normal and the
action--reaction pair --- the stated requirements --- are fully delivered, and
adding tangential impulses changes no data structures.

\textbf{Hand-rolled integration anywhere?} No. Free flight is SUNDIALS
exclusively; an impulse is an instantaneous momentum update at an event
boundary --- the standard treatment in every engine surveyed.

\end{document}
